Epistemic Semantics in Guarded String Models\thanks{Thanks guys.}

### 2021-01-14 tex/pdraft.tex
editing pass

\title{Epistemic Semantics in Guarded String Models\thanks{Thanks to Tobias Kappe for insightful reactions to the paper.
Thanks also to three SCiL reviewers for helpful comments.
Earlier versions of the work were presented in the LUSH talk series at Utrecht in December 2019, and in the Workshop on formal/philosophical/computational approaches to natural language meaning at Rochester in December 2020.  Thanks to the audiences for their reactions.
    .}}
\renewcommand\footnotemark{}
%%\renewcommand\footnoterule{}

@@

{\em Axiomatic Classification}. To situate our logic as a modal logic, consider the soundness of
the standard modal axioms given our semantics \citep{hughes1996new}. Some of
these standard axioms (see \ex{valid-axioms}) hold all the time, i.e. they
are \emph{valid}, and the remaining axioms are valid when $\hat{R}_a$
has a certain shape. The axioms in \ex{conditionally-valid-axioms}
have a nontrivial condition validity condition.

@@

\begin{example}\label{conditionally-valid-axioms}
  \(\begin{array}[t]{lll}
  \textbf{T} & \Box_a p \to p & \text{if } g\mathrel{\hat{R}_a} g, \forall g \\
  \textbf{D} & \Box_a p \to \Diamond_a p & \text{if } g \in \dom(\hat{R}_a), \forall g \\
  \textbf{4} & \Box_a p \to \Box_a\Box_a p & \text{if } \hat{R}_a \text{ idempotent}
  \end{array}\)
\end{example}

The condition on \textbf{4} is essentially just a restatement of the
theorem in relational terms. The conditions on the remaining standard
modal axioms, \textbf{B} and \textbf{5}, are of a similarly trivial
flavor.

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,linden2009hfst,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic \textsc{Kat} inside the finite state
calculus. We describe our translation from Epik logical terms to Fst terms
here.

@@

Defining \textsc{Kat} product using Fst's set-lifted string concatenation (denoted by
juxtaposition $\texttt{X}~\texttt{Y}$) requires more care. Naively concatenating
strings with atoms (Boolean vectors) at both ends doubles atoms at the juncture, and
does not enforce the requisite atom equality.  To implement \textsc{Kat} product, we define
the binary operation \texttt{Cn}, which concatenates strings in the string algebra,
removes strings with non-matching atoms, and then deletes the second of two atoms to
create a set of well-formed guarded strings.  See Figure \ref{fstfigure1}. {\tt Wf0}
is the set of ordinary strings that does not contain unequal pairs of atoms, as
defined using Fst's containment operator \texttt{\$}. The \texttt{Squash} relation
uses Fst's rewrite notation to delete atoms (elements of \texttt{St}) that are
preceded by another atom. \footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted in that the
  complement and set difference for non-equal length relations is not defined.
  Epistemic alterative relations are equal-length relations.} This relation is
applied via the relational composition (\texttt{.o.}) and codomain (\texttt{.l})
operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

\textsc{Kat} Kleene plus is defined in a similar way using Kleene plus in the string algebra,
with checks for equality of atoms and deletion of atoms. 
\textsc{Kat} Kleene star is defined from \textsc{Kat} Kleene plus and the multiplicative identity,
which is the set of well-formed atoms \texttt{St}.

@@

It remains to define an epistemic alternative relation on worlds for each agent.  The
relevant information in Figure \ref{epikfigure} is a relation between bare events for
each agent. This determines a relation between bare events decorated with compatible
atoms.  In Fst, we use the closure of the concatenation product operation on
relations to lift a relation on decorated events for an agent to the corresponding
relation on worlds.  The concatenation product $R~S$ of two relations $R$ and $S$ is
the set of pairs of the form $\pair{x_1x_2}{y_1y_2}$, where $x_1 \mathrel R y_1$, and
$x_2 \mathrel S y_2$.  In Fst, {\tt{R+}} is the closure of relation {\tt R} with
respect to this operation. Figure \ref{relclosure} defines the corresponding
operation on sets of guarded strings as encoded in Fst.\footnote{Relation
  concatenation in Fst differs from relation composition ({\tt{.o.}}), and the
  closure under discussion here is the closure of the former rather than the latter.}
The epistemic alternative relation on worlds for an agent is then defined as the \textsc{Kat}
relation concatenation closure {\tt RelKst} of the decorated-event alternative
relation for the agent.

@@

\section{Discussion}
Epistemic \textsc{Kat} is designed for use in research in linguistic semantics,
for research in computational linguistics on model-theoretically grounded
semantics, and for education at the level of a second graduate course in formal
semantics, covering intensionality. There are straightforward extensions to
additional linguistic phenomena, such as tense and perfective aspect
\exi{extension}{a}, and the combination of metaphysical modality and prospective
aspect \exi{extension}{b}.

@@

The model framework is a constructive branching-time framework with
metaphysical modality and epistemic modality, which will be applicable
in linguistic semantic research on combinations of tense, metaphysical
modality, and epistemic complementation
\citep{thomason1984combinations, abusch1998generalizing,
  condoravdi2002temporal}.  Connections with research on temporal
constitution of events in a related formal setting remain to be
explored
\citep{fernando2004finite,fernando2007observing,carlson2009tense}.

The syntactic part of the grammar formalism uses categorial grammar to
map between strings (or trees or derivations) and logical
translations.  This way of using a categorial grammar is standard
\citep{steedman2000syntactic,bozsahin2012combinatory}, and there are
computational implementations for it
\citep{barker2005parser,bozsahin2021ccglab}.\footnote{Our grammar uses
  basic categorial grammar, not additional features such as
  combinators, type raising, continuations, or multimodal slashes that
  interact with the grammar in non-trivial ways.  These features 
  become relevant in more sophisticated grammars.  Multimodal
  categories are used for phrases such as a tenseless verb phrase that
  semantically are functional, but combine in the grammar only as
  arguments.}  While this way of proceeding is simple and attractive
for the applications we are interested in, it would be possible to use
our semantics with other frameworks that have a semantics with a
lambda extension of a logical language.

>>>>

\title{Epistemic Semantics in Guarded String Models\thanks{Thanks to Tobias Kapp\'{e} for insightful reactions to the paper.
Thanks also to three SCiL reviewers for helpful comments.
Earlier versions of the work were presented in the LUSH talk series at Utrecht in December 2019, and in the Workshop on formal/philosophical/computational approaches to natural language meaning at Rochester in December 2020.  Thanks to the audiences for their reactions.}}
\renewcommand\footnotemark{}
%%\renewcommand\footnoterule{}

@@

{\em Axiomatic Classification}. To situate our logic as a modal logic, consider
the soundness of the standard modal axioms given our semantics
\citep{hughes1996new}. The axioms in \ex{valid-axioms}

@@

The axioms in \ex{conditionally-valid-axioms} are not valid, but do hold when
when the relation $R_a$ has a certain shape.

\begin{example}\label{conditionally-valid-axioms}
  \(\begin{array}[t]{lll}
  \textbf{T} & \Box_a p \to p & \text{if } g \mathrel{\hat{R}_a} g, \forall g \\
  \textbf{D} & \Box_a p \to \Diamond_a p & \text{if } \hat{R}_a \text{ is a function} \\
  \textbf{4} & \Box_a p \to \Box_a\Box_a p & \text{if } \hat{R}_a \text{ is idempotent}
  \end{array}\)
\end{example}

The condition on \textbf{4} is essentially just a restatement of the
theorem in relational terms. The conditions on the remaining standard
modal axioms, \textbf{B} and \textbf{5}, are similar.

\section{Translation into the finite state calculus}
The finite state calculus is an algebra of regular sets of strings and
regular relations between strings that was designed for use in
computationonal phonology and morphographemics
\citep{kaplan1994regular,beesley2003finite}.  Current implementations
allow for the definition of functions on regular sets and relations
\citep{hulden2009foma,linden2009hfst,karttunen2010update}.  Such definitions are used
here to construct of a model for epistemic \textsc{Kat} inside the finite state
calculus (Figures~\ref{fstfigure1}\&\ref{relclosure}).

@@

Defining \textsc{Kat} product using Fst's set-lifted string concatenation (denoted by
juxtaposition $\texttt{X}~\texttt{Y}$) requires more care. Naively concatenating
strings with atoms (Boolean vectors) at both ends doubles atoms at the juncture, and
does not enforce the requisite atom equality.  To implement \textsc{Kat} product, we define
the binary operation \texttt{Cn}, which concatenates strings in the string algebra,
removes strings with non-matching atoms, and then deletes the second of two atoms to
create a set of well-formed guarded strings (Figure \ref{fstfigure1}). {\tt Wf0}
is the set of ordinary strings with only equal pairs of atoms, as
defined using Fst's containment operator \texttt{\$}. The \texttt{Squash} relation
uses Fst's rewrite notation to delete atoms (elements of \texttt{St}) that are
preceded by another atom\footnote{ This is a non-equal length regular relation. The
  finite state calculus includes such relations, and they can be used with relation
  composition and relation domain and co-domain.  They are restricted in that the
  complement and set difference for non-equal length relations is not defined.
  Epistemic alterative relations are equal-length relations.}. This relation is
applied via the composition (\texttt{.o.}) and codomain (\texttt{.l})
operators.
%% Deletion is accomplished by a
%% re-write rule in the finite state calculus, which is a notation for
%% defining regular relations by contextually constrained substitutions.
%% In this case, is {\tt Squash} is a regular relation that deletes an
%% atom (a sequence of 0's and 1's of a certain length) when it follows
%% an atom.

Kleene plus is defined in a similar way, using Kleene plus in the string
algebra, with checks for equality of atoms and deletion of atoms. Kleene star is
defined from Kleene plus the well-formed atoms \texttt{St}, which implements $1$.

@@

Finally, the epistemic alternative relation for each agent, i.e. the
\texttt{agent} relation from Figure~\ref{epikfigure} is implemented in Fst.
Combining the agent relation with the effect formula gives a relation on events
decorated with compatible atoms. Then the relation on worlds is constructed
using the closure of the concatenation product operation. The concatenation
product $R~S$ of two relations $R$ and $S$ is the set of pairs of the form
$\pair{x_1x_2}{y_1y_2}$, where $x_1 \mathrel R y_1$, and $x_2 \mathrel S y_2$.
In Fst, {\tt{R+}} is the closure of relation {\tt R} with respect to this
operation. Figure \ref{relclosure} defines the corresponding operation on sets
of guarded strings as encoded in Fst.\footnote{Relation concatenation in Fst
  differs from relation composition ({\tt{.o.}}), and the closure under
  discussion here is the closure of the former rather than the latter.} The
epistemic alternative relation on worlds for an agent is then defined as the
\textsc{Kat} relation concatenation closure {\tt RelKst} of the decorated-event
alternative relation for the agent.

@@

\section{Discussion}
Epistemic \textsc{Kat} is designed for use in linguistic semantics research, for
compultational linguistics research on model-theoretically grounded semantics,
and for students in graduate course modules on intensional formal semantics.

There are straightforward extensions to additional linguistic phenomena, such as
tense and perfective aspect \exi{extension}{a}, and the combination of
metaphysical modality and prospective aspect \exi{extension}{b}.

@@

The model framework is a constructive branching-time framework with metaphysical
and epistemic modalities, which will be applicable in linguistic semantic
research on combinations of tense, metaphysical modality, and epistemic
complementation \citep{thomason1984combinations, abusch1998generalizing,
  condoravdi2002temporal}. Connections with research on temporal constitution of
events remain to be explored
\citep{fernando2004finite,fernando2007observing,carlson2009tense}.

The syntactic part of the grammar formalism uses a standard approach to
categorial grammar \citep{steedman2000syntactic,bozsahin2012combinatory} to map
between strings (or trees or derivations) and logical terms, and is rigorous
enough to admit computational implementations
\citep{barker2005parser,bozsahin2021ccglab}.\footnote{Our grammar uses basic
  categorial grammar, not additional features such as combinators, type raising,
  continuations, or multimodal slashes that interact with the grammar in
  non-trivial ways. These features become relevant in more sophisticated
  grammars. Multimodal categories are used for phrases such as a tenseless verb
  phrase that semantically are functional, but combine in the grammar only as
  arguments.} While this approach is simple and attractive for our applications,
it would be possible to use Epik with other frameworks that support lambda
extensions of logical languages.
